---
title: "COVID and GDP"
author: "casey durfee"
date: "1/18/2022"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rvest)
```

## Description

I decided to look at how COVID fatalities correlate with income in the United States.

I obtained county-level COVID data from Johns Hopkins University and the county-level 
per capita income data from Wikipedia.

For COVID, I am using the total death counts as of 12/31/2021.

The income data is from the 2009–2013 American Community Survey 5-Year Estimates.
```{r get_data}
covid_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"

deaths_by_county <- read_csv(covid_url, show_col_types = FALSE) %>%
      mutate(CountyState = paste(Admin2, Province_State, sep = ', ')) %>%
      rename("Deaths" = "12/31/21") %>%
      select(CountyState, Deaths) 

gdp_table <- read_html("https://en.wikipedia.org/wiki/List_of_United_States_counties_by_per_capita_income") %>%
        html_element(xpath = "//table[3]") %>%
        html_table()

colnames(gdp_table) <- c("Rank", "County", "State",
        "PerCapitaIncome", "MedianHousehold", "MedianFamily", "Population", "NumHouseholds")

gdp_table <- gdp_table %>%
              filter(Rank != "—") %>%
              mutate(CountyState = paste(str_trim(County), State, sep = ', '),
                    PerCapitaIncome = parse_number(PerCapitaIncome),
                    MedianHousehold = parse_number(MedianHousehold),
                    MedianFamily = parse_number(MedianFamily),
                    Population = parse_number(Population),
                    NumHouseholds = parse_number(NumHouseholds),
                    Rank = parse_number(Rank)
              )
```

We'll merge the two data samples together, and calculate the per-capita
fatality rate for COVID using the population counts from Wikipedia.

```{r merge}

merged <- left_join(deaths_by_county, gdp_table, by = c("CountyState" = "CountyState")) %>%
        mutate(PerCapitaDeaths = Deaths / Population) %>% 
        filter(!is.na(PerCapitaIncome))

cor(merged$PerCapitaDeaths, merged$PerCapitaIncome)
cor(merged$PerCapitaDeaths, merged$MedianHousehold)
cor(merged$PerCapitaDeaths, merged$MedianFamily)
```
All of the income measures are correlated with per-capita deaths, with median family 
income being the strongest predictor.

```{r linearmodel}
l <- lm(merged$PerCapitaDeaths ~ merged$MedianFamily)

summary(l)

plot(merged$MedianFamily, merged$PerCapitaDeaths,
     xlab="Median Family Income",
     ylab="Per Capita Deaths from COVID") + abline(l, col='red') 
```


## Missing Data
There are many data points reporting zero COVID
deaths. It turns out some of those counties are fairly large:

```{r zerocovid}
zero_deaths <- merged %>% filter(PerCapitaDeaths == 0)

zd <- zero_deaths[order(zero_deaths$Population, decreasing = TRUE), ]

head(zd)
```

It's highly unlikely that Weber County, with a population of 233,871, has had
zero COVID fatalities. Currently about 3/1000 Americans have died of COVID, so
we would expect about 700 COVID deaths in Weber County.

In total, there are about 700,000 people living in counties with zero reported 
fatalities in the Johns Hopkins data.  Utah is the worst offender.

```{r discrepancies}
sum(zd$Population)

by_state <- zd %>% group_by(State) %>% summarize(total = sum(Population))

by_state[order(by_state$total, decreasing = TRUE), ]
```

So, it's worth redoing the fit without the 0 fatality counties, even though 
there probably are some small counties that actually don't have any fatalities.

```{r}
without_zero_deaths <- merged %>%
                filter(PerCapitaDeaths != 0) 

l <- lm(without_zero_deaths$PerCapitaDeaths ~ without_zero_deaths$MedianFamily)

summary(l)

plot(without_zero_deaths$MedianFamily,
     without_zero_deaths$PerCapitaDeaths,
     xlab="Median Family Income",
     ylab="Per Capita Deaths from COVID") + abline(l, col='red')
```

The fit is only slightly better, with the R-squared value going from .19 to .20.

## Bias/Conclusion
There appears to be a correlation between income and COVID 
fatality rate, although the COVID data is incomplete from some
counties, particularly in the state of Utah.

I don't think I did anything biased in my analysis, but I did get the result 
I was expecting -- that richer counties would have a lower COVID fatality rate 
than poorer counties. 

There may also be other biases in the data -- over or under-reporting of COVID deaths, 
for instance. Additionally, the per capita income data is based on estimates from 
2013, which isn't ideal.

I think the next step would be to compare counties within states, and compare states
with similar states, in order to control for other factors and focus specifically on 
income.

In Colorado, for instance, do poorer counties have a higher COVID
fatality rate than richer counties? And how well does the state-level GDP predict 
COVID fatality rate? That would probably give more insight to the link between GDP
and COVID rather than comparing every county in the US with each other.

```{r sessionInfo}
sessionInfo()
```
